---
title: "Assignment 4 Version 4: Analysis and Visualisations"
author: "Rikhitha Varma"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

# Purpose of this version

This version develops the first analytical findings from the cleaned dataset.

# Setup

```{r}
library(tidyverse)
library(knitr)
library(scales)
```

# Load processed data

```{r}
analysis_data <- read_csv("processed_data/analysis_data.csv")

glimpse(analysis_data)
```

# Summary statistics

```{r}
analysis_data %>%
  summarise(
    average_distress = mean(distress_pct, na.rm = TRUE),
    average_low_life_satisfaction = mean(life_satisfaction_pct, na.rm = TRUE),
    average_food_insecurity = mean(food_insecurity_pct, na.rm = TRUE),
    average_irsd = mean(irsd_score, na.rm = TRUE)
  )
```

# LGAs with highest psychological distress

```{r}
analysis_data %>%
  arrange(desc(distress_pct)) %>%
  select(lga_name, distress_pct, food_insecurity_pct, irsd_score) %>%
  slice_head(n = 10) %>%
  kable(digits = 2)
```

# Food insecurity and psychological distress

```{r}
ggplot(analysis_data, aes(x = food_insecurity_pct, y = distress_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Food insecurity and psychological distress across Victorian LGAs",
    x = "Food insecurity (%)",
    y = "High or very high psychological distress (%)"
  )
```

# SEIFA disadvantage and psychological distress

```{r}
ggplot(analysis_data, aes(x = irsd_score, y = distress_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "SEIFA disadvantage and psychological distress across Victorian LGAs",
    x = "IRSD score",
    y = "High or very high psychological distress (%)"
  )
```

# Correlation tests

```{r}
cor.test(
  analysis_data$food_insecurity_pct,
  analysis_data$distress_pct
)
```

```{r}
cor.test(
  analysis_data$irsd_score,
  analysis_data$distress_pct
)
```

# Save figures

```{r}
dir.create("figures", showWarnings = FALSE)

food_plot <- ggplot(analysis_data, aes(x = food_insecurity_pct, y = distress_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Food insecurity and psychological distress",
    x = "Food insecurity (%)",
    y = "High or very high psychological distress (%)"
  )

ggsave(
  "figures/food_insecurity_distress.png",
  food_plot,
  width = 7,
  height = 5
)

irsd_plot <- ggplot(analysis_data, aes(x = irsd_score, y = distress_pct)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "SEIFA disadvantage and psychological distress",
    x = "IRSD score",
    y = "High or very high psychological distress (%)"
  )

ggsave(
  "figures/irsd_distress.png",
  irsd_plot,
  width = 7,
  height = 5
)
```

```{r}
ggplot(
  analysis_data,
  aes(
    x = life_satisfaction_pct,
    y = distress_pct
  )
) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Life satisfaction and psychological distress",
    x = "Low life satisfaction (%)",
    y = "High psychological distress (%)"
  )
```

```{r}
cor.test(
  analysis_data$life_satisfaction_pct,
  analysis_data$distress_pct
)
```

```{r}
analysis_data %>%
  arrange(distress_pct) %>%
  select(
    lga_name,
    distress_pct,
    life_satisfaction_pct,
    irsd_score
  ) %>%
  slice_head(n = 10) %>%
  knitr::kable(digits = 2)
```

```{r}
ggplot(
  analysis_data,
  aes(x = distress_pct)
) +
  geom_histogram(bins = 15) +
  labs(
    title = "Distribution of psychological distress across Victorian LGAs",
    x = "High or very high psychological distress (%)",
    y = "Number of LGAs"
  )
```

```{r}
library(sf)

lga_map <- st_read(
  "raw_data/boundaries/LGA_2021_AUST_GDA2020_SHP/LGA_2021_AUST_GDA2020.shp",
  quiet = TRUE
)

names(lga_map)
```

```{r}
victoria_map <- lga_map %>%
  filter(STE_NAME21 == "Victoria")

nrow(victoria_map)
```

```{r}
unique(victoria_map$STE_NAME21)
```

```{r}
head(victoria_map$LGA_NAME21, 20)
```

```{r}
victoria_map_clean <- victoria_map %>%
  mutate(
    lga_name = LGA_NAME21,
    lga_name = if_else(lga_name == "Bayside (Vic.)", "Bayside", lga_name)
  )

map_data <- victoria_map_clean %>%
  left_join(
    analysis_data,
    by = "lga_name"
  )

map_data %>%
  st_drop_geometry() %>%
  summarise(
    total_polygons = n(),
    matched = sum(!is.na(distress_pct)),
    unmatched = sum(is.na(distress_pct))
  )
```

```{r}
ggplot(map_data) +
  geom_sf(aes(fill = distress_pct), colour = "white", linewidth = 0.1) +
  labs(
    title = "Psychological distress across Victorian LGAs",
    fill = "Distress (%)"
  ) +
  theme_minimal()
```

```{r}
ggplot(map_data) +
  geom_sf(aes(fill = irsd_score), colour = "white", linewidth = 0.1) +
  labs(
    title = "SEIFA disadvantage across Victorian LGAs",
    fill = "IRSD score"
  ) +
  theme_minimal()
```

```{r}
analysis_data %>%
  arrange(desc(distress_pct)) %>%
  select(
    lga_name,
    distress_pct,
    irsd_score,
    food_insecurity_pct
  ) %>%
  slice_head(n = 5) %>%
  knitr::kable(
    caption = "Five LGAs with the highest psychological distress"
  )
```

```{r}
analysis_data %>%
  arrange(distress_pct) %>%
  select(
    lga_name,
    distress_pct,
    irsd_score,
    food_insecurity_pct
  ) %>%
  slice_head(n = 5) %>%
  knitr::kable(
    caption = "Five LGAs with the lowest psychological distress"
  )
```

```{r}
metro_lgas <- c(
  "Banyule","Bayside","Boroondara","Brimbank","Cardinia",
  "Casey","Darebin","Frankston","Glen Eira","Greater Dandenong",
  "Hobsons Bay","Hume","Kingston","Knox","Manningham",
  "Maribyrnong","Maroondah","Melbourne","Melton","Monash",
  "Moonee Valley","Moreland","Mornington Peninsula",
  "Nillumbik","Port Phillip","Stonnington","Whitehorse",
  "Whittlesea","Wyndham","Yarra","Yarra Ranges"
)
```

```{r}
analysis_data %>%
  mutate(
    region = ifelse(
      lga_name %in% metro_lgas,
      "Metropolitan",
      "Regional"
    )
  ) %>%
  group_by(region) %>%
  summarise(
    average_distress = mean(distress_pct, na.rm = TRUE),
    average_food_insecurity = mean(food_insecurity_pct, na.rm = TRUE),
    average_life_satisfaction = mean(life_satisfaction_pct, na.rm = TRUE)
  ) %>%
  knitr::kable(
    digits = 2,
    caption = "Average wellbeing indicators for metropolitan and regional LGAs"
  )
```

# 

# Changes from Version 3

## Summary of Changes

This version was compared to Version 3, which focused on exploratory analysis, statistical testing and spatial investigation of mental health and wellbeing patterns across Victorian LGAs.

The following changes were made:

### Data validation and inspection

Before beginning the analysis, the processed dataset was reloaded from `analysis_data.csv` to ensure that all previous data preparation steps had been completed correctly. The dataset was checked to confirm that it contained 75 Victorian LGAs and the expected variables required for the analysis. Variable types were reviewed and missing values were identified to ensure the data was suitable for statistical analysis and visualisation. These checks helped verify that the processed dataset accurately reflected the outputs generated during the data cleaning and integration stage.

### Descriptive statistics

To gain an initial understanding of the data, summary statistics were calculated for psychological distress, low life satisfaction, food insecurity and socioeconomic disadvantage (IRSD). Statewide averages were produced to provide a baseline for comparison across LGAs. In addition, a histogram was created to examine how psychological distress was distributed across Victoria and to identify whether some LGAs experienced substantially higher or lower distress levels than others.

### Identification of LGAs with highest and lowest psychological distress

To better understand geographic variation in mental wellbeing, Victorian LGAs were ranked according to their levels of psychological distress. Tables were produced showing the five LGAs with the highest levels of distress and the five LGAs with the lowest levels. Food insecurity and IRSD scores were included alongside these results to provide additional socioeconomic context and to explore whether highly distressed communities shared similar characteristics.

### Exploratory visualisations

A series of scatterplots were created to visually investigate potential relationships between psychological distress and other wellbeing indicators. These included food insecurity, socioeconomic disadvantage (IRSD), and low life satisfaction. Regression trend lines were added to each plot to assist interpretation and highlight any underlying patterns. A histogram was also produced to visualise the overall distribution of psychological distress across Victorian LGAs.

### Statistical analysis

Pearson correlation tests were conducted to quantify the relationships observed in the visualisations. Three separate relationships were examined: food insecurity and psychological distress, IRSD score and psychological distress, and low life satisfaction and psychological distress. Both the strength of the relationship and its statistical significance were assessed to determine whether any observed associations were likely to be meaningful.

### Spatial analysis

To explore geographic patterns in wellbeing outcomes, Victorian LGA boundary shapefiles were loaded using the `sf` package and joined to the analysis dataset. The join process was checked to ensure that wellbeing indicators were correctly matched to their corresponding LGAs. Choropleth maps were then created to visualise the spatial distribution of psychological distress and socioeconomic disadvantage across Victoria. These maps helped identify regional patterns that may not have been apparent through tables or statistical summaries alone.

### Metropolitan and regional comparison

As a final analytical step, LGAs were grouped into metropolitan and regional categories. Average levels of psychological distress, food insecurity and low life satisfaction were calculated for each group. Comparing these averages provided an additional perspective on how wellbeing outcomes differ between metropolitan and regional Victoria and helped place the earlier findings into a broader geographic context.

## Interpretation of Results

### Relationship between food insecurity and psychological distress

A scatterplot and Pearson correlation analysis were used to investigate the association between food insecurity and psychological distress. It was found that there was almost no correlation between the two variables (r = -0.004, p = 0.974). No pattern was observed in a scatter plot and the fitted regression line was very nearly horizontal.

This indicates that there are no direct links at LGA level between food insecurity and psychological distress. While the relationship between food security and social determinants of health is well acknowledged, the outcomes of food insecurity could also be more apparent at the household and/or individual level than at the aggregate LGA level analysed in this study..

### Relationship between socioeconomic disadvantage and psychological distress

\
The Index of Relative Socio-economic Disadvantage (IRSD) was used to examine the relationship between socioeconomic disadvantage and psychological distress. Any negative IRSD scores suggest disadvantage, any positive scores suggest relative advantage.

There was only a weak negative correlation, as indicated by the scatterplot. Pearson correlation analysis was used to determine if there was statistically significant relationship between IRSD score with psychological distress (r = -0.061, p = 0.606).

The finding suggests that the socioeconomic disadvantage factor is not the only factor accounting for the inter-LGA scales in psychological distress in Victoria. The results indicate that mental wellbeing is related to a wider mix of economic, social and environmental determinants that are not necessarily measured at an area level.

### Relationship between life satisfaction and psychological distress

The strongest relationship seen in the analysis was for low life satisfaction.

The scatter plot indicated that there was a clear positive association among LGA indicators of low life satisfaction with the indicators of psychological distress. Pearson correlation analysis revealed the high and statistically significant correlation (r = 0.764, p \< 0.001).

Such a discovery indicates that subjective wellbeing measures can be a better predictor of mental health outcomes of the population than their socioeconomic status. The degree of dissatisfaction with life in the community was significantly associated with the level of psychological distress reported by residents.The community's level of dissatisfaction with life was significantly correlated with residents' levels of psychological distress.

### Spatial patterns in mental wellbeing

The spatial analysis showed significant geographical variation in psychological distress across Victoria. Distress levels over 30% were noted in several LGAs; distress levels under 10% were noted in other LGAs.

Darebin, Whittlesea, Moreland, Knox and Port Phillip had the highest levels of psychological distress. South Gippsland, Queenscliffe, Buloke, Southern Grampians and Swan Hill were the lowest.

Interestingly, both high and low distress LGA had a wide spread of IRSD scores. There was some psychological distress in relatively high-SES LGAs and some low psychological distress in relatively low-SES LGAs. These results provide further evidence for the hypothesis that socioeconomic disadvantage is an inadequate explanation of differences in mental wellbeing outcomes.

The maps also illustrated that important local level variations can be masked by the statewide averages, underscoring the usefulness of spatial analysis in the investigation of public health outcomes.

### Metropolitan and regional Victoria

There were significant differences between wellbeing outcomes between metropolitan and regional LGAs.

The average psychological distress rate was found to be higher for metropolitan LGAs (23.6%) compared to regional LGAs (19.1%). Low life satisfaction was reported at higher rates in the metropolitan areas as well. The food insecurity among the two groups was fairly comparable.

The findings indicate that psychological distress is more likely to be occurring in metropolitan Victoria despite the overall high levels of access to services and economic opportunities. Other factors like housing affordability stress, population density, social isolation, cost of living stress may all play a role in these differences.

## Key Findings

The analysis produced five key findings:

1.  Psychological distress varies substantially across Victorian LGAs, ranging from less than 10% to more than 30%.

2.  There was no meaningful relationship between food insecurity and psychological distress across Victorian LGAs (r = -0.004).

3.  There was no statistically significant relationship between socioeconomic disadvantage (IRSD) and psychological distress (r = -0.061).

4.  Low life satisfaction showed a strong positive relationship with psychological distress (r = 0.764, p \< 0.001).

5.  Metropolitan LGAs reported higher average levels of psychological distress than regional LGAs.

Overall, the results suggest that subjective wellbeing measures, especially life satisfaction, are more closely associated with psychological distress than area-level measures of socioeconomic disadvantage or food insecurity. Results also point to considerable geographical variation in mental wellbeing outcomes across Victoria, and to the importance of local level analysis in the investigation of population health patterns.

Version 4 greatly improved the analytical aspect of the project with descriptive statistics, correlation testing, spatial visualisation and comparisons between metropolitan and regional areas. These results will be used directly to inform the final blog post and conclusions presented in the final submission.
